---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r, include = FALSE}
knitr::opts_chunk$set(collapse  = TRUE,
                      comment   = "#>",
                      fig.path  = "figures/readme/",
                      out.width = "100%")
```



StatLearn_FaPaCa
=========================================================



<p align="left">
  • <a href="#overview">Overview</a><br>
  • <a href="#data-sources">Data sources</a><br>
  • <a href="#workflow">Workflow</a><br>
  • <a href="#content">Content</a><br>
  • <a href="#installation">Installation</a><br>
  • <a href="#usage">Usage</a><br>
  • <a href="#citation">Citation</a>
</p>



## Overview

This repository is the research compendium for **Proteomics biomarker discovery
for individualized prevention of familial pancreatic cancer using statistical
learning**. It contains the R code for two parts of the work:

- a **simulation study** comparing penalized and boosted regression models for
  binary outcomes with many correlated predictors, and
- an **analysis of the FaPaCa proteomics data** using the same methods.

The methods implemented are ridge regression, adaptive lasso, component-wise
boosting (`glmboost`, `gamboost` from `mboost`) and complementary pairs
stability selection (`stabs`) for adaptive lasso and boosting.

The workflows are defined as [`targets`](https://docs.ropensci.org/targets/)
pipelines that run in parallel with `crew`.



## Data sources

- **Simulated data** are generated by `get_simulated_data()`
  ([`R/data_preparation.R`](R/data_preparation.R)), with the sample size `N`,
  number of predictors `P`, number of informative predictors `p_ref`,
  intercept `tau` and correlation `rho` as parameters.
- **FaPaCa proteomics data** are not distributed with this repository. To run
  the FaPaCa analysis, place the raw data at `data/raw_data.csv`.



## Workflow

**Simulation study** ([`analyses/simulation_study_pipeline.R`](analyses/simulation_study_pipeline.R))

1. *Experiment 1*: simulate one data set, split it into stratified
   cross-validation training and test sets, fit ridge, adaptive lasso and
   glmboost on the training set, predict the test set, and extract the selected
   features (adaptive lasso) and variable importance (glmboost).
2. *Experiment 2*: simulate data while varying one parameter at a time (`N`,
   `P`, `p_ref`, `tau`, `rho`) and run stability selection (`q = 10`,
   `PFER = 2`) with adaptive lasso and glmboost.

**FaPaCa study** ([`analyses/fapaca_study.R`](analyses/fapaca_study.R))

1. Read the raw data and track the file for changes.
2. Impute missing values with `missForest`, then export and reload the imputed
   data.
3. Prepare the cross-validation and training data.
4. Fit glmboost and gamboost models.



## Content

This repository is structured as follows:

- [`DESCRIPTION`](DESCRIPTION):
contains project metadata (authors, date, dependencies, etc.)

- [`make.R`](make.R):
restores the package environment, loads the project functions and runs both pipelines

- [`R/`](R):
contains R functions developed especially for this project
  - [`data_preparation.R`](R/data_preparation.R): data simulation, cross-validation splits, reading, imputing and preparing the FaPaCa data
  - [`models.R`](R/models.R): model fitting (ridge, adaptive lasso, glmboost, gamboost), feature extraction and test-set prediction
  - [`stability_selection.R`](R/stability_selection.R): stability selection for adaptive lasso, glmboost and gamboost
  - [`auxiliary_functions.R`](R/auxiliary_functions.R): helper functions

- [`analyses/`](analyses):
the two `targets` pipelines (simulation study and FaPaCa study)

- [`outputs/`](outputs) and `_targets/`:
results produced by the pipelines

- [`data/`](data):
place for the raw data (not included)

- [`manuscript/`](manuscript) and [`figures/`](figures):
the published paper and its figures

- [`tests/`](tests):
unit tests

- [`Dockerfile`](Dockerfile), [`rproject.toml`](rproject.toml) and
[`rv.lock`](rv.lock):
computational environment (Docker image, [`rv`](https://github.com/A2-ai/rv)
dependency file and lock file)



## Installation

To install this compendium:

- [Fork](https://docs.github.com/en/get-started/quickstart/contributing-to-projects) 
this repository using the GitHub interface.
- [Clone](https://docs.github.com/en/repositories/creating-and-managing-repositories/cloning-a-repository) 
your fork using `git clone fork-url` (replace `fork-url` by the URL of your fork). 
Alternatively, open [RStudio IDE](https://posit.co/products/open-source/rstudio/) 
and create a New Project from Version Control.
- Install [`rv`](https://github.com/A2-ai/rv), then install the R packages
listed in `rproject.toml` with `rv sync` (or by running `source("make.R")`).

Alternatively, build the Docker image (based on `rocker/rstudio`) and open
RStudio at <http://localhost:8787> (user `rstudio`):

```bash
docker build -t statlearn-fapaca .
docker run -e PASSWORD=<choose-a-password> -p 8787:8787 statlearn-fapaca
```



## Usage

Install the packages, load the project functions and run both pipelines
with [`make.R`](make.R):

```{r eval=FALSE}
source("make.R")
```

The FaPaCa study is skipped when `data/raw_data.csv` is missing. To run a
single pipeline instead:

```{r eval=FALSE}
# Simulation study
targets::tar_make(script = "analyses/simulation_study_pipeline.R",
                  store = "outputs/simulation_study")

# FaPaCa study (requires data/raw_data.csv)
targets::tar_make(script = "analyses/fapaca_study.R",
                  store = "outputs/fapaca_study")
```

**Notes**

- Each pipeline keeps its `targets` store in `outputs/`
- Both pipelines run on 4 local `crew` workers
- Experiment 2 of the simulation study (stability selection) takes a long time
- `analyses/fapaca_study.R` still uses placeholder feature sets (`age`, `sex`)
that must be replaced with the proteomics features before use



## Citation

Please use the following citation: 

> Ha CSR, Müller-Nurasyid M, Petrera A, Hauck SM, Marini F, Bartsch DK, Slater EP, Strauch K.
> Proteomics biomarker discovery for individualized prevention of familial pancreatic cancer using statistical learning.
> *PLoS One*. 2023 Jan 26;18(1):e0280399.
> doi: [10.1371/journal.pone.0280399](https://doi.org/10.1371/journal.pone.0280399).
> PMID: [36701413](https://pubmed.ncbi.nlm.nih.gov/36701413/);
> PMCID: [PMC9879447](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9879447/).
